---
title: "Collective defense strategies"
editor_options: 
  chunk_output_type: console
---

# Isolates selection

## Rationale

The first phase of this study aimed to **select an entomopathogenic fungal isolate** to evaluate how a toxic substrate affects the collective immune defense of *Atta sexdens*.

Pathogenicity was evaluated on *A. sexdens* foragers exposed to five *Metarhizium robertsii* isolates—kindly provided by the *Laboratório de Interações Inseto-Microrganismos* at the *Universidade Federal de Viçosa*.

Twenty-four hours prior to the bioassays, colonies were starved. Following starvation, **colonies were randomly assigned to two groups and exposed to one of two substrate regimes for seven days**: (i) 10.0 g of *Acalypha wilkesiana* leaves daily (control group), or (ii) 10.0 g of fresh *Tropaeolum majus* leaves daily, serving as a potential modulating agent of social defenses.

After this period, **individual ants from these colonies were immersed for two seconds in a conidial suspension** at a concentration of 1x10⁶ conidia/mL, containing distilled water and 0.01% Tween 80. **Three control groups were evaluated**: (1) an aqueous solution of 0.01% Tween 80 (vehicle control), (2) distilled water (negative control), and (3) a 0.0001% isocycloseram solution (positive control). Each treatment consisted of four replicates (Petri dishes) with five foragers each. Ants were housed in Petri dishes lined with moistened filter paper and maintained in a BOD incubator at 27 ± 2 °C, 70 ± 10% relative humidity, and continuous darkness (0:24 h L:D). The ants were starved throughout the **4-day evaluation period**, and mortality was recorded daily. Dead individuals were immediately removed and transferred to individual moist chambers (microcentrifuge tubes containing cotton and autoclaved distilled water) to evaluate fungal sporulation (confirming mycosis).

## Statistical analysis

The statistical analysis for the pathogenicity bioassay data was conducted in R using a multi-step framework. Survival curves were generated using the Kaplan–Meier estimator and compared across isolates and controls via the log-rank test to evaluate daily mortality rates. To quantify the hazard ratios (HR) of substrate type, fungal treatment, and their interaction (substrate:treatment), a mixed-effects Cox proportional hazards model (`coxme`) was fitted, specifying the Petri dish (replicate) as a random effect to account for pseudo-replication. Finally, the median lethal time (LT₅₀) for each treatment combination was estimated using dose-response modeling via the `drc` package.

## Results

Instalando e carregando os pacotes

```{r}
library(readxl) #leitura do arquivo excel
library(dplyr) #manipulação dos dados
library(survival) #análise de sobrevivência
library(survminer) #gráficos de sobrevivência
#install.packages("coxme")
library(coxme) #modelo de cox com efeitos aleatórios
library(lme4) #modelos lineares generalizados mistos (glmm)
library(car) #ANOVA tipo III para glmm
#install.packages("emmeans")
library(emmeans) #comparações mútiplas post-hoc
#install.packages()
library(drc) #curvas dose-resposta e cálculo de TL50
library(tidyverse)
```

Lendo e preparando os dados

```{r}
dados <- read_excel("isolates_selection_bioassay_1.xlsx")

#definindo as variáveis categóricas e criando um ID único por placa de Petri
dados <- dados %>%
  mutate(
    substrate = as.factor(substrate),
    treatment = as.factor(treatment),
    inter     = as.factor(inter),
    rep       = as.factor(rep),
    id_placa  = as.factor(paste(inter, rep, sep = "_"))
  )
```

### Survival curves (Kaplan-Meier)

```{r}
#criando o objeto de sobrevivência
surv_obj <- Surv(time = dados$dias, event = dados$mortality)

#curvas de sobrevivência por tratamento de interação
fit_km <- survfit(surv_obj ~ inter, data = dados)

#teste de Log-Rank
logrank_test <- survdiff(surv_obj ~ inter, data = dados)
print(logrank_test)

#plotando a curva de sobrevivência
ggsurvplot(
  fit_km,
  data = dados,
  pval = TRUE,
  risk.table = FALSE,
  legend.title = "Treatments",
  xlab = "Days",
  ylab = "Survival probability",
  ggtheme = theme_minimal()
)
```

### Mixed-effects Cox model (main effects + interaction)

```{r}
fit_cox <- coxme(surv_obj ~ substrate * treatment + (1 | id_placa), data = dados)
summary(fit_cox)
```

### Median lethal time (LT₅₀) estimation

```{r}
lt50_matrix <- quantile(fit_km, probs = 0.5)$quantile

lt50_table <- data.frame(
  Treatment_Group = gsub("inter=", "", rownames(lt50_matrix)),
  LT50_days = lt50_matrix[, 1],
  row.names = NULL
)

print(lt50_table)
```

Fungal pathogenicity and substrate interaction significantly affected Atta sexdens forager survival over time (Log-rank χ2 = 76.6, df = 15, p \< 0.001). The mixed-effects Cox proportional hazards model revealed a significant interaction between substrate type and fungal isolate on mortality hazard (p = 0.0122). Specifically, exposure to isolate i94 on Tropaeolum majus substrate increased mortality risk 4.7-fold (HR = 4.70, z = 2.51, p = 0.0122) relative to control conditions. On the control substrate, isolate i33 displayed the highest independent toxicity (HR = 2.58, z = 2.28, p = 0.0223). Overall mortality was high across treatments and approached saturation by day 5, supporting the Cox proportional hazards analysis as the primary approach for assessing differences in mortality risk among treatments.
