---
title: "Collective defense strategies"
editor_options: 
  chunk_output_type: console
---

# Isolates selection

## Rationale

The first phase of this study aimed to **select an entomopathogenic fungal isolate** to evaluate how a toxic substrate affects the collective immune defense of *Atta sexdens*.

Pathogenicity was evaluated on *A. sexdens* foragers exposed to five *Metarhizium robertsii* isolates—kindly provided by the *Laboratório de Interações Inseto-Microrganismos* at the *Universidade Federal de Viçosa*.

Twenty-four hours prior to the bioassays, colonies were starved. Following starvation, **colonies were randomly assigned to two groups and exposed to one of two substrate regimes for seven days**: (i) 10.0 g of *Acalypha wilkesiana* leaves daily (control group), or (ii) 10.0 g of fresh *Tropaeolum majus* leaves daily, serving as a potential modulating agent of social defenses.

After this period, **individual ants from these colonies were immersed for two seconds in a conidial suspension** at a concentration of 1x10⁶ conidia/mL, containing distilled water and 0.01% Tween 80. **Three control groups were evaluated**: (1) an aqueous solution of 0.01% Tween 80 (vehicle control), (2) distilled water (negative control), and (3) a 0.0001% isocycloseram solution (positive control). Each treatment consisted of four replicates (Petri dishes) with five foragers each. Ants were housed in Petri dishes lined with moistened filter paper and maintained in a BOD incubator at 27 ± 2 °C, 70 ± 10% relative humidity, and continuous darkness (0:24 h L:D). The ants were starved throughout the **4-day evaluation period**, and mortality was recorded daily. Dead individuals were immediately removed and transferred to individual moist chambers (microcentrifuge tubes containing cotton and autoclaved distilled water) to evaluate fungal sporulation (confirming mycosis).

## Statistical analysis

The statistical analysis for the pathogenicity bioassay data was conducted in R using a multi-step framework. Survival curves were generated using the Kaplan–Meier estimator and compared across isolates and controls via the log-rank test to evaluate daily mortality rates. To quantify the hazard ratios (HR) of substrate type, fungal treatment, and their interaction (substrate:treatment), a mixed-effects Cox proportional hazards model (`coxme`) was fitted, specifying the Petri dish (replicate) as a random effect to account for pseudo-replication. Finally, the median lethal time (LT₅₀) for each treatment combination was estimated using dose-response modeling via the `drc` package.

## Results

Instalando e carregando os pacotes

```{r}
library(readxl) #leitura do arquivo excel
library(dplyr) #manipulação dos dados
library(survival) #análise de sobrevivência
library(survminer) #gráficos de sobrevivência
#install.packages("coxme")
library(coxme) #modelo de cox com efeitos aleatórios
library(lme4) #modelos lineares generalizados mistos (glmm)
library(car) #ANOVA tipo III para glmm
#install.packages("emmeans")
library(emmeans) #comparações mútiplas post-hoc
#install.packages()
library(drc) #curvas dose-resposta e cálculo de TL50
library(tidyverse)
```

### Foraging data

```{r}
foraging_exp1 <- read_excel("foraging_bioassay_1.xlsx")

ggplot(data = foraging_exp1, aes(x = real_consume, fill = substrate)) +
    geom_histogram(alpha = .4) +
    labs(title = "Histogram")

ggplot(data = foraging_exp1, aes(x = day, y = real_consume, fill = substrate)) +
    geom_point(shape = 21, size = 4, color = "black")

ggplot(data = foraging_exp1, aes(x = day, y = real_consume,
                            color = substrate, shape = substrate)) +
    geom_point(size = 4, alpha = .7) +
    geom_smooth(method = "lm", se = FALSE) +
    scale_shape_manual(values = c(15, 17)) +
    scale_color_manual(values = c("purple", "cyan4")) +
    theme_bw(base_size = 16) +
    labs(x = "day", y = "Substrate consume (g)", 
         color = "substrate", shape = "substrate")
```

Para analisar dados de consumo diário com medições repetidas nas mesmas colônias ao longo de 7 dias, a abordagem ideal é utilizar um Modelo Linear Misto (LMM) para isolar a autocorrelação temporal das colônias e contornar o baixo poder estatístico inicial do tamanho amostral (n = 3 colônias por substrato).Estatística Descritiva e ExploratóriaMétricas Diárias: Calcule média, erro-padrão da média (EPM), mediana e amplitude interquartil de real_consume agrupados por substrate e day.Objetivo: Mapear a trajetória central do consumo ao longo do tempo e verificar a variabilidade inter-colonial de cada grupo, além de identificar eventuais outliers.Análise Inferencial: Modelo Linear Misto (LMM)Por que usar: Cada colônia foi medida 7 vezes consecutivas, violando a premissa de independência de uma ANOVA tradicional. O LMM insere colony como efeito aleatório $(1 \vert{} \text{colony})$ para controlar a variação individual, enquanto substrate e day entram como efeitos fixos testando a interação $\text{substrate} \times \text{day}$.Validação de Premissas: Avalie a normalidade dos resíduos do modelo via teste de Shapiro-Wilk. Caso os resíduos violem a normalidade, aplica-se transformação nos dados (ex: $\log(y + 1)$) ou utiliza-se um Modelo Linear Generalizado Misto (GLMM) com distribuição Gama.Estratégia de Visualização CientíficaPara relatórios ou publicações, o gráfico ideal combina a tendência central do tratamento com a variação individual de cada unidade experimental (spaghetti plot sobreposto por médias):Linhas finas e transparentes ao fundo: Trajetória individual diária das 6 colônias (mostra a variação biológica real).Linhas espessas em destaque: Média diária de cada substrato ao longo dos 7 dias.Sombreamento (Ribbon): Representação gráfica do Erro-Padrão da Média ($\pm \text{SE}$).

```{r}
# 1. Pacotes Necessários --------------------------------------------------
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, lme4, lmerTest, car, emmeans)

# 2. Dados Exemplo (Substitua pela leitura do seu CSV) -------------------
# dados <- read.csv("seu_arquivo.csv")

set.seed(42)
dados <- expand.grid(day = 1:7, colony = factor(1:6), substrate = c("Substrato_A", "Substrato_B")) %>%
  filter((substrate == "Substrato_A" & colony %in% 1:3) | 
         (substrate == "Substrato_B" & colony %in% 4:6)) %>%
  mutate(real_consume = rnorm(n(), mean = ifelse(substrate == "Substrato_A", 10 + day*1.5, 8 + day*0.4), sd = 1.2))

# 3. Estatística Descritiva -----------------------------------------------
descritiva <- dados %>%
  group_by(substrate, day) %>%
  summarise(
    n = n(),
    media = mean(real_consume, na.rm = TRUE),
    se = sd(real_consume, na.rm = TRUE) / sqrt(n),
    mediana = median(real_consume, na.rm = TRUE),
    .groups = "drop"
  )

print("--- RESUMO DESCRITIVO ---")
print(descritiva)

# 4. Modelo Linear Misto (LMM) -------------------------------------------
dados$day_num <- as.numeric(dados$day)

# Modelo testando efeito fixo do substrato, dia e interação, com efeito aleatório da colônia
modelo <- lmer(real_consume ~ substrate * day_num + (1 | colony), data = dados)

# Teste de Normalidade dos Resíduos
print("--- NORMALIDADE DOS RESÍDUOS ---")
print(shapiro.test(residuals(modelo)))

# Tabela ANOVA (Tipo III)
print("--- TABELA ANOVA (EFEITOS FIXOS) ---")
print(Anova(modelo, type = "III"))

# Médias Ajustadas e Comparações
print("--- MÉDIAS AJUSTADAS (EMMEANS) ---")
print(emmeans(modelo, ~ substrate | day_num))

# 5. Visualização Científica (ggplot2) -----------------------------------
grafico <- ggplot() +
  # Trajetórias individuais de cada colônia
  geom_line(data = dados, 
            aes(x = day, y = real_consume, group = colony, color = substrate),
            alpha = 0.35, linetype = "dashed", linewidth = 0.7) +
  # Intervalo de erro-padrão
  geom_ribbon(data = descritiva, 
              aes(x = day, ymin = media - se, ymax = media + se, fill = substrate),
              alpha = 0.2) +
  # Linha média do substrato
  geom_line(data = descritiva, 
            aes(x = day, y = media, color = substrate, group = substrate),
            linewidth = 1.2) +
  geom_point(data = descritiva, 
             aes(x = day, y = media, color = substrate), 
             size = 3) +
  # Estética e Escalas
  scale_x_continuous(breaks = 1:7) +
  scale_color_brewer(palette = "Set1", name = "Substrato") +
  scale_fill_brewer(palette = "Set1", name = "Substrato") +
  labs(
    x = "Tempo de Exposição (Dias)",
    y = "Consumo Real (g)",
    title = "Dinâmica Diária de Forrageamento",
    subtitle = "Média ± EPM (n = 3 colônias/substrato). Linhas pontilhadas indicam colônias individuais."
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

print(grafico)
```

### Survival curves (Kaplan-Meier)

Lendo e preparando os dados

```{r}
dados <- read_excel("isolates_selection_bioassay_1.xlsx")

#definindo as variáveis categóricas e criando um ID único por placa de Petri
dados <- dados %>%
  mutate(
    substrate = as.factor(substrate),
    treatment = as.factor(treatment),
    inter     = as.factor(inter),
    rep       = as.factor(rep),
    id_placa  = as.factor(paste(inter, rep, sep = "_"))
  )
```
```{r}
#criando o objeto de sobrevivência
surv_obj <- Surv(time = dados$dias, event = dados$mortality)

#curvas de sobrevivência por tratamento de interação
fit_km <- survfit(surv_obj ~ inter, data = dados)

#teste de Log-Rank
logrank_test <- survdiff(surv_obj ~ inter, data = dados)
print(logrank_test)

#plotando a curva de sobrevivência
ggsurvplot(
  fit_km,
  data = dados,
  pval = TRUE,
  risk.table = FALSE,
  legend.title = "Treatments",
  xlab = "Days",
  ylab = "Survival probability",
  ggtheme = theme_minimal()
)
```

### Mixed-effects Cox model (main effects + interaction)

```{r}
fit_cox <- coxme(surv_obj ~ substrate * treatment + (1 | id_placa), data = dados)
summary(fit_cox)
```

### Median lethal time (LT₅₀) estimation

```{r}
lt50_matrix <- quantile(fit_km, probs = 0.5)$quantile

lt50_table <- data.frame(
  Treatment_Group = gsub("inter=", "", rownames(lt50_matrix)),
  LT50_days = lt50_matrix[, 1],
  row.names = NULL
)

print(lt50_table)
```

Fungal pathogenicity and substrate interaction significantly affected Atta sexdens forager survival over time (Log-rank χ2 = 76.6, df = 15, p \< 0.001). The mixed-effects Cox proportional hazards model revealed a significant interaction between substrate type and fungal isolate on mortality hazard (p = 0.0122). Specifically, exposure to isolate i94 on Tropaeolum majus substrate increased mortality risk 4.7-fold (HR = 4.70, z = 2.51, p = 0.0122) relative to control conditions. On the control substrate, isolate i33 displayed the highest independent toxicity (HR = 2.58, z = 2.28, p = 0.0223). Overall mortality was high across treatments and approached saturation by day 5, supporting the Cox proportional hazards analysis as the primary approach for assessing differences in mortality risk among treatments.
